---
title: "IFR Flows"
author: "Enrico Spinielli"
date: "`r Sys.Date()`"
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "220px",
  fig.align = 'center'
)

library(dplyr)
library(stringr)
library(countrycode)
library(readr)
library(magrittr)
library(ifrflows)
```


# Abstract

This document describes the assumptions, shortcuts and steps taken to prepare
the data to generate a circular plot depicting the flow of flights
entering / exiting the airspace of Eurocontrol member States for the years 2007-2015.


# Introduction

The visualization part of this work is inspired by Sander, Abel and Bauer's website [see @sander_2014] and its
relevant paper [see @abel_2014a].

The data is collected from the PRU's Analytical database and includes all 2001-2015 IFR flights,
but the analysis below will consider only the interval 2007-2015.


# The nasty details

## The Flights

The query used to export the relevant flows in CSV format to the file `ifr_flows_2001_2015.csv` the same as
encapsulated in the `retrieve_country_flows` function:

```{r yearly-country-flows, eval=FALSE}
retrieve_country_flows
```

The flollowing filters have been applied:

* flights without an ISO2 country code for either the departure or destination airport have been excluded:
    
    ```{sql connection=NULL, eval=FALSE}
      ...
      AND (A1.EC_ISO_CT_CODE != '##' AND A2.EC_ISO_CT_CODE != '##')
      ...
    
    ```


* military and helicopter flights have been excluded:

    
    ```{sql connection=NULL, eval=FALSE}
    ...
    AND (F.ICAO_FLT_TYPE != 'M')
    AND SUBSTR(AC.ICAO_DESC,1,1) != 'H'
    ...
    ```

The amount of filtered out flights is tiny compared with the relevant flows:

* **TODO** table with counts for no ISO code flights
* **TODO** table for military and heli flights


## Data preparation for the visualization library

The data as pre-prepard in the CSV file `ifr_flows_2001-2015.csv` is the result of exporting
what retrieved by running `retrieve_country_flows('2001', '2015')`.
The relevant dataset can be loaded as follows:

```{r read-raw}
raw_flows_file = "ifr_flows_2001-2015.csv"
extra_flows_file = "extra-flows.csv"
intra_flows_file = "intra-flows.csv"

# explicitly mentrion `na` option and exclude the default 'NA' 
# because 'NA' is a valid ISO code for NAMIBIA and not a missing value
cols <- cols(
  YEAR = col_integer(),
  NB_OF_FLIGHT = col_integer(),
  DEP_ISO_COUNTRY_CODE = col_character(),
  DES_ISO_COUNTRY_CODE = col_character()
)

flows <- read_csv(
  system.file("extdata", raw_flows_file, package = "ifrflows"),
  na = c(""),
  col_types = cols)
flows
```


**Note**: the ISO code for Serbia is provided as CS, which was correct till 2005.
Since 2006 it is RS.
Let's fix it globally (even if we should keep `CS` before 2006...):
Also `XF` has been used for Martinique/Guadaloupe, let's put `MQ` which will result in a proper UN region.

```{r}
flows <- flows %>%
  fix_iso_countrycode()
```


Now remove the rows where ISO2 code is NA (for 2014 it is Namibia only which accounts for in/out flows of less than 400 flights per year...)

```{r}
# visually inspect, should be 0 given that NAMIBIA, NA is valid ISO
flows %>%
  filter(is.na(DEP_ISO_COUNTRY_CODE) | is.na(DES_ISO_COUNTRY_CODE)) %>%
  group_by(DEP_ISO_COUNTRY_CODE,DES_ISO_COUNTRY_CODE) %>%
  summarize(filtered_out = sum(NB_OF_FLIGHT)) %>%
  print(n=100)

flows %>%
  filter(is.na(DEP_ISO_COUNTRY_CODE)) %>%
  summarise(filtered_out_dep = sum(NB_OF_FLIGHT))

flows %>%
  filter(is.na(DES_ISO_COUNTRY_CODE)) %>%
  summarise(filtered_out_des = sum(NB_OF_FLIGHT))

# filter out the entry where there is no ISO code
flows <- flows %>%
  filter(!(is.na(DEP_ISO_COUNTRY_CODE) | is.na(DES_ISO_COUNTRY_CODE)))
```

Rename the variables
```{r}
flows <- flows %>%
  rename(
    year = YEAR,
    flights = NB_OF_FLIGHT,
    origin_iso = DEP_ISO_COUNTRY_CODE,
    destination_iso = DES_ISO_COUNTRY_CODE)
```




## Defining regions

Let's get the UN country codes.
Luckily there is an R package for that, and I have contributed to it in order to define custom regions for PRU
(and STATFOR).



## Matching country with region


```{r}
flows <- flows %>%
  mutate(origin_region_name = countrycode(origin_iso, "iso2c", "region"),
         destination_region_name = countrycode(destination_iso, "iso2c", "region"),
         origin_name = countrycode(origin_iso, "iso2c", "country.name"),
         destination_name = countrycode(destination_iso, "iso2c", "country.name"))

flows <- flows %>%
  select(year, flights,
         origin_iso, origin_name, origin_region_name,
         destination_iso, destination_name, destination_region_name)
```


### Check for `AN`

There is a problem with `AN`, the former Netherland Antilles. Since 2011 `AN` has been split in three codes:
`BQ` (Bonaire, with Bonaire International Airport: (IATA: BON, ICAO: TNCB)),
`CW` (Curaçao, with Curaçao Internationa Airport: (IATA: CUR, ICAO: TNCC) and
`SX` (Sint Eustatius, with F. D. Roosevelt Airport (IATA: EUX, ICAO: TNCE). This is a very small airport).

In our dataset we have also `TNCM` (Princess Juliana International Airport (IATA: SXM, ICAO: TNCM))
with countrycode `AN`.


```{r an-fix}
flows <- flows %>%
  fix_region_AN()
```



## Check for `TW`

Certain ISO 2 alpha have officially no UN region, i.e. Taiwan and British Indian Ocean Territory.
Let's fix it:

```{r}
flows <- flows %>%
  fix_region_TW() %>% 
  fix_region_IO()
```


## Format for circular plot

Define regions as per Eurocontrol classification

```{r}
classification <- "eurocontrol_pru"
# handle 'AN', 'TW' and 'IO' and 'SN'
flows <- flows %>%
  mutate(origin_region_name = countrycode(origin_iso, "iso2c", classification),
         destination_region_name = countrycode(destination_iso, "iso2c", classification),
         origin_name = countrycode(origin_iso, "iso2c", "country.name"),
         destination_name = countrycode(destination_iso, "iso2c", "country.name")) %>% 
  fix_eurocontrol_region()
```


We will now split flows into intra and extra-European flows:

* intra flows: flights departing **AND** arriving to EUROCONTROL region
* extra flows:

    - flights departing from EUROCONTROL region **AND**
      arriving to another region but EUROCONTROL
    - flights departing from another region but EUROCONTROL **AND**
      arriving to EUROCONTROL region

We will exclude overflying flights, i.e. the ones departing **AND** arriving from
another region but EUROCONTROL. (TODO: assess amounts...)

```{r}
# flights that stay within EUROCONTROL Member States
intra_flows <- flows %>%
  filter(origin_region_name == "Eurocontrol" & destination_region_name == "Eurocontrol")


# flows from/to outside EUROCONTROL Member States
extra_flows <- flows %>%
  filter(
    # flowing IN to EUROCONTROL region
    (origin_region_name != "Eurocontrol" & destination_region_name == "Eurocontrol") | 
    # flowing OUT of EUROCONTROL region
    (origin_region_name == "Eurocontrol" & destination_region_name != "Eurocontrol"))
```


### Extra flows

Relabel / regroup regions

```{r}
# Russia ->  Asia
extra_flows <- extra_flows %>%
  fix_extra_flows_region()
```


Let's analyse the size of the flows per year

```{r, eval=FALSE}
extra_stats <- extra_flows %>%
  mutate(flow_name = ifelse(
    origin_region_name < destination_region_name,
    paste0(origin_region_name, "-", destination_region_name),
    paste0(destination_region_name, "-", origin_region_name)
  )) %>%
  select(year, flights, flow_name) %>%
  group_by(year, flow_name) %>%
  summarise(movements = sum(flights)) %>%
  arrange(desc(movements)) %>%
  spread(year, movements, fill = 0)
```

We then need a file that specifies which countries to show, something like:

```csv
iso,show (1 = yes; 2 = no)
US,1
IN,1
CN,2
...
```

In fact which countries to filer out is dictated by them having a too small flows.
Say we keep only the flows with at least `threshold` flights

```{r}
require(tidyr)

# list of unfiltered countries
extra_destination_countries <- extra_flows %>%
  select(destination_iso, destination_name) %>%
  rename(iso = destination_iso, name = destination_name) %>%
  distinct()

extra_departure_countries <- extra_flows %>%
  select(origin_iso, origin_name) %>%
  rename(iso = origin_iso, name = origin_name) %>%
  distinct()

# set visibility to FALSE (0)
extra_countries <- dplyr::union(extra_destination_countries, extra_departure_countries) %>%
  distinct() %>%
  arrange(iso) %>%
  mutate(show = 0)

# Regions in the (unfiltered) set
extra_destination_regions <- extra_flows %>%
  select(destination_region_name) %>%
  rename(name = destination_region_name) %>%
  distinct()

extra_departure_regions <- extra_flows %>%
  select(origin_region_name) %>%
  rename(name = origin_region_name) %>%
  distinct()

extra_regions <- dplyr::union(extra_destination_regions, extra_departure_regions) %>%
  dplyr::distinct() %>%
  arrange(name)
```

Now let's filter out the little flows...

```{r}
extra_years <- extra_flows %>%
  select(year) %>%
  distinct() %>%
  arrange(year)

# threshold for filtering out small flows 
extra_fpd <- 10 # flights per day
extra_threshold <- 365 * extra_fpd

# eventually keep only a subset of the time serie
# extra_years <- extra_years[[1]] # get the array of _ALL_ year numbers
# years <- c(years[length(years)]) # just take last one
extra_years_subset <- extra_years %>% 
  pull(year)

# keep only the years as previously defined, used to generate the list of visible countries
extra_flows <- extra_flows %>% 
  filter(flights > extra_threshold) %>%
  filter(year %in% extra_years_subset) %>%
  spread(year, flights, fill = 0)

# prepend "countryflow_"" to year columns, i.e. from "2007" to "countryflow_2007"
names(extra_flows) <- ifelse(names(extra_flows) %in% unlist(extra_years, use.names = FALSE),
                       paste("countryflow_", names(extra_flows), sep = ""),
                       names(extra_flows))
```



### Intra flows

For the flows inside EUROCONTROL region, use STATFOR regions:

```{r}
# split intra-Eurocontrol flights in regions according to STATFOR
classification <- "eurocontrol_statfor"
intra_flows <- intra_flows %>%
  mutate(origin_region_name = countrycode(origin_iso, "iso2c", classification),
         destination_region_name = countrycode(destination_iso, "iso2c", classification),
         origin_name = countrycode(origin_iso, "iso2c", "country.name"),
         destination_name = countrycode(destination_iso, "iso2c", "country.name"))
```




```{r}
# do the same for intra
intra_years <- intra_flows %>%
  select(year) %>%
  distinct() %>%
  arrange(year)

names(intra_years) <- c("year")

intra_years_subset <- intra_years %>% 
  pull(year)

# threshold for filtering out small flows 
intra_fpd <- 0 # flights per day
intra_threshold <- 365 * intra_fpd

intra_flows <- intra_flows %>% 
  filter(flights > intra_threshold) %>%
  filter(year %in% intra_years_subset) %>%
  spread(year, flights, fill = 0)

# prepend "countryflow_"" to year columns, i.e. from "2007" to "countryflow_2007"
names(intra_flows) <- ifelse(names(intra_flows) %in% unlist(intra_years, use.names=FALSE),
                       paste("countryflow_", names(intra_flows), sep=""),
                       names(intra_flows))
```

Now let's create the files for the countries and the regions present in the filtered dataset

```{r}
intra_destination_countries <- intra_flows %>%
  select(destination_iso, destination_name) %>%
  rename(iso = destination_iso, name = destination_name) %>%
  distinct()

intra_departure_countries <- intra_flows %>%
  select(origin_iso, origin_name) %>%
  rename(iso = origin_iso, name = origin_name) %>%
  dplyr::distinct()

intra_countries <- dplyr::union(intra_destination_countries, intra_departure_countries) %>%
  dplyr::distinct() %>%
  arrange(iso) %>%
  mutate(show = 1)


# Regions in the (unfiltered) set
intra_destination_regions <- intra_flows %>%
  select(destination_region_name) %>%
  rename(name = destination_region_name) %>%
  distinct()

intra_departure_regions <- intra_flows %>%
  select(origin_region_name) %>%
  rename(name = origin_region_name) %>%
  distinct()

intra_regions <- dplyr::union(intra_destination_regions, intra_departure_regions) %>%
  dplyr::distinct() %>%
  arrange(name)
```

List of countries contributing the major movements

```{r}
# limit to 2014 (for now)
intra_movements_2014 <- intra_flows %>% 
  select(origin_name, destination_name,
         origin_region_name,
         destination_region_name,
         countryflow_2014) %>%
  mutate(flow = ifelse(origin_region_name != 'Eurocontrol',
                       origin_name,
                       destination_name))

# print 30 biggest contributors
intra_movements_2014 %>% 
  group_by(flow) %>% 
  summarise(total = sum(countryflow_2014)) %>%
  arrange(desc(total)) %>%
  print(n = 30)
```


**TODO**: Analyse the distribution of flows


### Save to files

Save all on file...

```{r save-files, eval=FALSE}
data_dir <- str_glue("{root}/data/", root = here())
flows %>%
  write_csv(path = paste(data_dir, "ifr_flows_2014-2015.csv", sep = ""))

# INTRA
intra_flows %>%
  write_csv(path = paste(data_dir, intra_flows_file, sep = ""))
intra_countries %>%
  write_csv(path = paste(data_dir, "intra_countries.csv", sep = ""))

intra_regions %>%
  write_csv(path = paste(data_dir, "intra_regions.csv", sep = ""))

intra_years %>%
  write_csv(path = paste(data_dir, "intra_years.csv", sep = ""))



# EXTRA
extra_flows %>%
  write_csv(path = paste(data_dir, extra_flows_file, sep = ""))

# quotes needed due to commas in some country names
extra_countries %>%
  write_csv(path = paste(data_dir, "extra_countries.csv", sep = ""))

extra_regions %>%
  write_csv(path = paste(data_dir, "extra_regions.csv", sep = ""))

extra_years %>%
  write_csv(path = paste(data_dir, "extra_years.csv", sep = ""))
```



## Plot
Sankey

```{r sankey, out.width="100%"}
library("dplyr")
library(tidygraph)
library(igraph)
library("networkD3")

cols <- cols(
  .default = col_integer(),
  origin_iso = col_character(),
  origin_name = col_character(),
  origin_region_name = col_character(),
  destination_iso = col_character(),
  destination_name = col_character(),
  destination_region_name = col_character()
)

intra <- read_csv(
  system.file("extdata", "ifr_flows_2014-2015.csv", package = "ifrflows"),
  na = c(""),
  col_types = cols)

flights_per_day <- 40
threshold <- 365 * flights_per_day

ii <- intra  %>%
  filter(year == 2015) %>% 
  select(origin_iso, destination_iso, flights) %>%
  dplyr::filter(flights > threshold) %>% 
  mutate(
    origin_iso = str_c(origin_iso, "dep"),
    destination_iso = str_c(destination_iso, "des")
  )


ng <- ii %>% 
  as_tbl_graph() %>% 
  igraph_to_networkD3()

sankeyNetwork(
  Links = ng[["links"]], 
  Nodes = ng[["nodes"]],
  Source = "source",
  Target = "target" ,
  Value = "value",
  NodeID = "name",
  fontSize = 14
)
```



# References
