---
title: "IFR Flows"
author: "Enrico Spinielli"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, results='asis', echo=FALSE, include=FALSE}
suppressPackageStartupMessages(library(here))
library(tidyverse)
library(countrycode)
library(readr)
library(magrittr)


knitr::read_chunk(str_glue('{root}/R/db.R', root = here()))
```


# Abstract

This document describes the assumptions, shortcuts and steps taken to prepare
the data to generate a circular plot depicting the flow of flights
entering / exiting the airspace of Eurocontrol member States for the years 2007-2015.


# Introduction

The visualization part of this work is inspired by Sander, Abel and Bauer's website [see @sander_2014] and its
relevant paper [see @abel_2014a].

The data is collected from the PRU's Analytical database and includes all 2001-2015 IFR flights, but the analysis below will consider only the interval 2007-2015.


# The nasty details

## The Flights

The query used to export the relevant flows in CSV format to the file `ifr_flows_2001_2015.csv` the same as
encapsulated in the `retrieve_country_flows` function:

```{r yearly-country-flows, eval=FALSE}
<<retrieve_country_flows>>
```

The flollowing filters have been applied:

* flights without an ISO2 country code for either the departing or arriving airport have been excluded:

  ```sql
  ...
  AND (A1.EC_ISO_CT_CODE != '##' AND A2.EC_ISO_CT_CODE != '##')
  ...
  ```

* military and helicopter flights have been excluded:


  ```sql
  ...
  AND (F.ICAO_FLT_TYPE != 'M')
  AND SUBSTR(AC.ICAO_DESC,1,1) != 'H'
  ...
  ```

The amount of filtered out flights is tiny compared with the relevant flows:

* **TODO** table with counts for no ISO code flights
* **TODO** table for military and heli filghts


## Data preparation for the visualization library

The data as pre-prepard in the CSV file `ifr_flows_2001-2015.csv` is the result of exporting
what retrieved by running `retrieve_country_flows('2001', '2015')`.
The relevant dataset can be loaded as follows:

```{r read-raw}
raw_flows_file = "ifr_flows_2001-2015.csv"
extra_flows_file = "extra-flows.csv"
intra_flows_file = "intra-flows.csv"

# explicitly mentrion `na` option and exclude the default 'NA' 
# because 'NA' is a valid ISO code for NAMIBIA and not a missing value
cols <- cols(
  YEAR = col_integer(),
  NB_OF_FLIGHT = col_integer(),
  DEP_ISO_COUNTRY_CODE = col_character(),
  DES_ISO_COUNTRY_CODE = col_character()
)

flows <- read_csv(str_glue('{root}/data/', raw_flows_file, root = here()), na = c(""), col_types = cols)
```


**Note**: the ISO code for Serbia is provided as CS, which was correct till 2005.
Since 2006 it is RS.
Let's fix it globally (even if we should keep `CS` before 2006...):

```{r}
flows %<>%
  mutate(
    DEP_ISO_COUNTRY_CODE = ifelse(DEP_ISO_COUNTRY_CODE == "CS", "RS", DEP_ISO_COUNTRY_CODE),
    DES_ISO_COUNTRY_CODE = ifelse(DES_ISO_COUNTRY_CODE == "CS", "RS", DES_ISO_COUNTRY_CODE))
```

Also `XF` has been used for Martinique/Guadaloupe, let's put `MQ` which will result in a proper UN region.
```{r}
flows %<>%
  mutate(
    DEP_ISO_COUNTRY_CODE = ifelse(DEP_ISO_COUNTRY_CODE == "XF", "MQ", DEP_ISO_COUNTRY_CODE),
    DES_ISO_COUNTRY_CODE = ifelse(DES_ISO_COUNTRY_CODE == "XF", "MQ", DES_ISO_COUNTRY_CODE))
```

Now remove the rows where ISO2 code is NA (for 2014 it is Namibia only which accounts for in/out flows of less than 400 flights per year...)

```{r}
# visually inspect, should be 0 given that NAMIBIA, NA is valid ISO
flows %>%
  filter(is.na(DEP_ISO_COUNTRY_CODE) | is.na(DES_ISO_COUNTRY_CODE)) %>%
  group_by(DEP_ISO_COUNTRY_CODE,DES_ISO_COUNTRY_CODE) %>%
  summarize(filtered_out = sum(NB_OF_FLIGHT)) %>%
  print(n=100)

flows %>%
  filter(is.na(DEP_ISO_COUNTRY_CODE)) %>%
  summarise(filtered_out_dep = sum(NB_OF_FLIGHT))

flows %>%
  filter(is.na(DES_ISO_COUNTRY_CODE)) %>%
  summarise(filtered_out_des = sum(NB_OF_FLIGHT))

# filter out the entry where there is no ISO code
flows %<>%
  filter(!(is.na(DEP_ISO_COUNTRY_CODE) | is.na(DES_ISO_COUNTRY_CODE)))
```

Rename the variables
```{r}
flows %<>%
  rename(
    year = YEAR,
    flights = NB_OF_FLIGHT,
    origin_iso = DEP_ISO_COUNTRY_CODE,
    destination_iso = DES_ISO_COUNTRY_CODE)
```




## Defining regions

Let's get the UN country codes.
Luckily there is an R package for that, and I have contributed to it in order to define custom regions for PRU
(and STATFOR).



## Matching country with region


```{r}
flows %<>%
  mutate(originregion_name = countrycode(origin_iso, "iso2c", "region"),
         destinationregion_name = countrycode(destination_iso, "iso2c", "region"),
         origin_name = countrycode(origin_iso, "iso2c", "country.name"),
         destination_name = countrycode(destination_iso, "iso2c", "country.name"))

flows %<>%
  select(year, flights,
         origin_iso, origin_name, originregion_name,
         destination_iso, destination_name, destinationregion_name)
```


### Check for `AN`

There is a problem with `AN`, the former Netherland Antilles. Since 2011 `AN` has been split in three codes:
`BQ` (Bonaire, with Bonaire International Airport: (IATA: BON, ICAO: TNCB)),
`CW` (Curaçao, with Curaçao Internationa Airport: (IATA: CUR, ICAO: TNCC) and
`SX` (Sint Eustatius, with F. D. Roosevelt Airport (IATA: EUX, ICAO: TNCE). This is a very small airport).

In our dataset we have also `TNCM` (Princess Juliana International Airport (IATA: SXM, ICAO: TNCM))
with countrycode `AN`.


```{r an-fix}
flows %<>%
  mutate(originregion_name = ifelse(origin_iso == "AN",
                                    "Caribbean",
                                    originregion_name)) %>%
  mutate(destinationregion_name = ifelse(destination_iso == "AN",
                                         "Caribbean",
                                         destinationregion_name))

```



## Check for `TW`

But certain ISO 2 alpha have officially no UN region, i.e. Taiwan and British Indian Ocean Territory.
Let's fix it:

```{r}
flows %<>%
  mutate(originregion_name = ifelse(origin_iso == "TW",
                                    "Eastern Asia",
                                    originregion_name)) %>%
  mutate(destinationregion_name = ifelse(destination_iso == "TW",
                                         "Eastern Asia",
                                         destinationregion_name)) %>% 
  mutate(originregion_name = ifelse(origin_iso == "IO",
                                    "Eastern Africa",
                                    originregion_name)) %>%
  mutate(destinationregion_name = ifelse(destination_iso == "IO",
                                         "Eastern Africa",
                                         destinationregion_name))
```


## Format for circular plot

Define regions as per Eurocontrol classification

```{r}
classification <- "eurocontrol_pru"
# handle 'AN', 'TW' and 'IO' and 'SN'
flows %<>%
  mutate(originregion_name = countrycode(origin_iso, "iso2c", classification),
         destinationregion_name = countrycode(destination_iso, "iso2c", classification),
         origin_name = countrycode(origin_iso, "iso2c", "country.name"),
         destination_name = countrycode(destination_iso, "iso2c", "country.name")) %>% 
  mutate(originregion_name = ifelse(origin_iso == "AN",
                                    "Southern America",
                                    originregion_name)) %>%
  mutate(destinationregion_name = ifelse(destination_iso == "AN",
                                         "Southern America",
                                         destinationregion_name)) %>% 
  mutate(origin_name = ifelse(origin_iso == "AN",
                                    "Antilles",
                                    origin_name)) %>%
  mutate(destination_name = ifelse(destination_iso == "AN",
                                         "Antilles",
                                         destination_name)) %>% 
  mutate(originregion_name = ifelse(origin_iso == "TW",
                                    "Eastern Asia",
                                    originregion_name)) %>%
  mutate(destinationregion_name = ifelse(destination_iso == "TW",
                                         "Eastern Asia",
                                         destinationregion_name)) %>% 
  mutate(originregion_name = ifelse(origin_iso == "IO",
                                    "Eastern Africa",
                                    originregion_name)) %>%
  mutate(destinationregion_name = ifelse(destination_iso == "IO",
                                         "Eastern Africa",
                                         destinationregion_name)) %>% 
  mutate(originregion_name = ifelse(origin_iso == "SN",
                                    "Western Africa",
                                    originregion_name)) %>%
  mutate(destinationregion_name = ifelse(destination_iso == "SN",
                                         "Western Africa",
                                         destinationregion_name))

```


Split flows into intra and extra-European flows   

```{r}
intra_flows <- flows %>%
  filter(originregion_name == "Eurocontrol" & destinationregion_name == "Eurocontrol")


# let's try to filter out internal Eurocontrol traffic
extra_flows <- flows %>%
  filter(!(originregion_name == "Eurocontrol" & destinationregion_name == "Eurocontrol"))
```


```{r}
# split intra-Eurocontrol flights in regions according to STATFOR
classification <- "eurocontrol_statfor"
intra_flows %<>%
  mutate(originregion_name = countrycode(origin_iso, "iso2c", classification),
         destinationregion_name = countrycode(destination_iso, "iso2c", classification),
         origin_name = countrycode(origin_iso, "iso2c", "country.name"),
         destination_name = countrycode(destination_iso, "iso2c", "country.name"))

```


Concentrate on extra-European flows   

```{r}

flows <- extra_flows
```

Relabel / regroup regions

```{r}
# Russia ->  Asia
flows %<>%
  mutate(originregion_name = ifelse(originregion_name == "Russia",
                                    "Asia",
                                    originregion_name))
flows %<>%
  mutate(destinationregion_name = ifelse(destinationregion_name == "Russia",
                                         "Asia",
                                         destinationregion_name))


# China ->  Asia
flows %<>%
  mutate(originregion_name = ifelse(originregion_name == "China",
                                    "Asia",
                                    originregion_name))
flows %<>%
  mutate(destinationregion_name = ifelse(destinationregion_name == "China",
                                         "Asia",
                                         destinationregion_name))


# Southern America ->  America
flows %<>%
  mutate(originregion_name = ifelse(originregion_name == "Southern America",
                                    "America",
                                    originregion_name))
flows %<>%
  mutate(destinationregion_name = ifelse(destinationregion_name == "Southern America",
                                         "America",
                                         destinationregion_name))

# Northern America -> N Ame
flows %<>%
  mutate(originregion_name = ifelse(originregion_name == "Northern America",
                                    "America",
                                    originregion_name))
flows %<>%
  mutate(destinationregion_name = ifelse(destinationregion_name == "Northern America",
                                         "America",
                                         destinationregion_name))
```


Let's analyse the size of the flows per year

```{r, eval=FALSE}
stats <- flows %>%
  mutate(flow_name = ifelse(originregion_name < destinationregion_name, 
                            paste0(originregion_name, "-", destinationregion_name),
                            paste0(destinationregion_name, "-", originregion_name))) %>%
  select(year, flights, flow_name) %>%
  group_by(year, flow_name) %>%
  summarise(movements=sum(flights)) %>%
  arrange(desc(movements)) %>%
  spread(year, movements, fill = 0)
```

We then need a file that specifies which countries to show, something like:

```csv
iso,show (1 = yes; 2 = no)
US,1
IN,1
CN,2
...
```

In fact which countries to filer out is dictated by them having a too small flows.
Say we keep only the flows with at least `threshold` flights

```{r}
require(tidyr)

# list of unfiltered countries
descou <- flows %>%
  select(destination_iso, destination_name) %>%
  rename(iso = destination_iso, name = destination_name) %>%
  distinct()

depcou <- flows %>%
  select(origin_iso, origin_name) %>%
  rename(iso = origin_iso, name = origin_name) %>%
  distinct()

# set visibility to FALSE (0)
cous0 <- dplyr::union(descou, depcou) %>%
  distinct() %>%
  arrange(iso) %>%
  mutate(show = 0)



```

Now let's filter out the little flows...

```{r}
years <- flows %>%
  select(year) %>%
  distinct() %>%
  arrange(year)

# names(years) <- c("year")

# threshold for filtering out small flows (3650 = flow of 10 flights/day between dep/des country)
threshold <- 3650 / 2

# eventually keep only a subset of the time serie
years <- years[[1]] # get the array of _ALL_ year numbers
# years <- c(years[length(years)]) # just take last one

# keep only the years as previously defined, used to generate the list of visible countries
flows %<>% 
  filter(flights > threshold) %>%
  filter(year %in% years) %>%
  spread(year, flights, fill=0)

# prepend "countryflow_"" to year columns, i.e. from "2007" to "countryflow_2007"
names(flows) <- ifelse(names(flows) %in% unlist(years, use.names=FALSE),
                       paste("countryflow_", names(flows), sep=""),
                       names(flows))

```

```{r}
# do the same for intra
# keep only the years as previously defined, used to generate the list of visible countries
threshold <- 0
intra_flows %<>% 
  filter(flights > threshold) %>%
  filter(year %in% years) %>%
  spread(year, flights, fill=0)

# prepend "countryflow_"" to year columns, i.e. from "2007" to "countryflow_2007"
names(intra_flows) <- ifelse(names(intra_flows) %in% unlist(years, use.names=FALSE),
                       paste("countryflow_", names(intra_flows), sep=""),
                       names(intra_flows))
```

Now let's create the files for the countries and the regions present in the filtered dataset

```{r}
descou <- flows %>%
  select(destination_iso, destination_name) %>%
  rename(iso = destination_iso, name = destination_name) %>%
  distinct()

depcou <- flows %>%
  select(origin_iso, origin_name) %>%
  rename(iso = origin_iso, name = origin_name) %>%
  dplyr::distinct()

cous <- dplyr::union(descou, depcou) %>%
  dplyr::distinct() %>%
  arrange(iso) %>%
  mutate(show = 1)

# set show=1 to cous0 rows matching cous iso 
cous0 %<>%
  left_join(cous, by="iso") %>%
  mutate(show=show.y,name=name.x) %>%
  select(iso,name,show) %>%
  mutate_all(funs(replace(., which(is.na(.)), 0)))


# Regions in the (unfiltered) set
desreg <- flows %>%
  select(destinationregion_name) %>%
  rename(name = destinationregion_name) %>%
  distinct()

depreg <- flows %>%
  select(originregion_name) %>%
  rename(name = originregion_name) %>%
  distinct()

regs <- dplyr::union(desreg, depreg) %>%
  dplyr::distinct() %>%
  arrange(name)

```

List of countries contributing the major movements

```{r}
# limit to 2014 (for now)
movements <- flows %>% 
  filter(originregion_name == 'Eurocontrol' |
           destinationregion_name == 'Eurocontrol') %>%
  select(origin_name, destination_name,
         originregion_name,
         destinationregion_name,
         countryflow_2014) %>%
  mutate(flow = ifelse(originregion_name != 'Eurocontrol',
                       origin_name,
                       destination_name))

# print 30 biggest contributors
movements %>% 
  group_by(flow) %>% 
  summarise(total = sum(countryflow_2014)) %>%
  arrange(desc(total)) %>%
  print(n=30)
```


**TODO**: Analyse the distribution of flows



Save all on file
```{r}
data_dir <- "data/"
flows %>%
  write_csv(., paste(data_dir, extra_flows_file, sep = ""))

intra_flows %>%
  write_csv(., paste(data_dir, intra_flows_file, sep = ""))

# quotes needed due to commas in some country names
cous %>%
  write_csv(., file = paste(data_dir, "countries.csv", sep = ""), row.names = FALSE, quote = TRUE)

regs %>%
  write_csv(., file = paste(data_dir, "regions.csv", sep = ""), row.names = FALSE, quote = TRUE)

years %>%
  write_csv(., sep = ",", file = paste(data_dir, "years.csv", sep = ""),
    row.names = FALSE, col.names = c("year"), quote = TRUE)
```



## Plot
Sankey

```{r sankey}
library("tidyverse")
cols <- cols(
  .default = col_integer(),
  origin_iso = col_character(),
  origin_name = col_character(),
  originregion_name = col_character(),
  destination_iso = col_character(),
  destination_name = col_character(),
  destinationregion_name = col_character()
)
intra <- read_csv("data/intra-flows.csv", col_types = cols)
ii <- intra  %>%
  select(origin_iso, destination_iso, countryflow_2015) %>%
  rename(flights = countryflow_2015) %>%
  dplyr::filter(flights > 36500)

library(tidygraph)
library(igraph)
ng <- ii %>% 
  as_tbl_graph() %>% 
  igraph_to_networkD3()

library("networkD3")
sankeyNetwork(
  Links = ng[["links"]], 
  Nodes = ng[["nodes"]],
  Source = "source",
  Target = "target" ,
  Value = "value",
  NodeID = "name",
  fontSize = 14
)


```

# References
